### 2020-12-04 paper/localpsf.tex
editing intro
put in kriging stuff from overleaf document

If $\mathcal{A}$ were translation invariant (that is, if $A(y+h, x+h) = A(y,x)$ for all $x$, $y$, $h$) then it is straightforward to show that $\mathcal{A}$ would be a convolution operator, and the convolution kernel, $\varphi_x$, would be given by
\begin{equation}
\label{eq:convolution_kernel}
	\varphi_x(y) := A(y+x,x).
\end{equation}
The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source at some fixed point $x$, then translating the resulting impulse response function re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. Locally translation invariant operators may therefore be approximated by spatially varying weighted sums of convolution operators of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\mathcal{A}u\right)(v) \approx \left(\widetilde{\mathcal{A}}u\right)(v) := \left\langle v, \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right\rangle_{L^2(\Omega)},
\end{equation}
here $f \cdot g$ denotes pointwise multiplication, and $f \ast g$ denotes convolution of the functions $f$ and $g$. Here
\begin{itemize}
	\item $\varphi_i := \varphi_{x_i}$ is given by formula \eqref{eq:convolution_kernel}, but with $x$ replaced by $x_i$.
	\item $\{x_i\}_{i=1}^r \subset \Omega$ is a collection of points scattered throughout the domain.
	\item The functions $w_i(x)$ are spatially varying weighting function that interpolate the convolution kernels $\varphi_i$.
\end{itemize}
Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. The more locally translation invariant an operator is, the smaller the number of terms that are required in \eqref{eq:product_convolution} to achieve an accurate approximation.

@@

For background on product-convolution approximations, we recommend reading the following papers: [CITES].

@@

\subsection{Motivating example: PDE constrained inverse problems with highly informative data}
\label{sec:PDE_hessian_motivation}

@@

\section{Interpolation of impulse response samples}

>>>>

If $\mathcal{A}$ were translation invariant (that is, if $A(y+h, x+h) = A(y,x)$ for all $x$, $y$, $h$) then it is straightforward to show that $\mathcal{A}$ would be a convolution operator, and the convolution kernel, $\varphi_x$, would be given by
\begin{equation}
\label{eq:convolution_kernel}
	\varphi_x(y) := A(y+x,x).
\end{equation}
The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source at some fixed point $x$, then translating the resulting impulse response function re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. Locally translation invariant operators may therefore be approximated by spatially varying weighted sums of convolution operators of the form
\begin{equation}
\label{eq:product_convolution}
	\left(\mathcal{A}u\right)(v) \approx \left(\widetilde{\mathcal{A}}u\right)(v) := \left\langle v,~ \sum_{i=1}^r \varphi_i \ast \left(w_i \cdot u\right) \right\rangle_{L^2(\Omega)},
\end{equation}
here $f \cdot g$ denotes pointwise multiplication, and $f \ast g$ denotes convolution of the functions $f$ and $g$. Here
\begin{itemize}
	\item $\varphi_i := \varphi_{x_i}$ is given by formula \eqref{eq:convolution_kernel}, but with $x$ replaced by $x_i$.
	\item $\{x_i\}_{i=1}^r \subset \Omega$ is a collection of points scattered throughout the domain.
	\item The functions $w_i(x)$ are spatially varying weighting function that interpolate the convolution kernels $\varphi_i$.
\end{itemize}
Approximations of the form \eqref{eq:product_convolution} are known as \emph{product-convolution} approximations, because the action of each term in the sum consists of a pointwise product, followed by a convolution. The more locally translation invariant an operator is, the smaller the number of terms that are required in \eqref{eq:product_convolution} to achieve an accurate approximation.

@@

For background on product-convolution approximations, we recommend reading the following papers: [CITES]. For background on H-matrices, we recommend reading the following papers: [CITES].

@@

\subsection{Motivation: PDE constrained inverse problems with highly informative data}
\label{sec:PDE_hessian_motivation}

@@

\section{Interpolating impulse response samples}

\section{Poisson PSF interpolation}

We form an interpolant, $\widetilde{\varphi}_x$, of the point spread function (PSF), $\varphi_x$, as follows:
\begin{equation*}
	\varphi_x(y) \approx \widetilde{\varphi}_x(y) := \sum_{i=1}^q w_i(x)\varphi_{x_i}(y).
\end{equation*}
The functions $y \mapsto \varphi_{x_i}(y)$, are samples of the PSF at the sample points $x_i$, $i=1,\dots,q$. The functions $w_i(x)$, are smooth spatially varying weighting functions that solve the following optimization problem:
\begin{equation}
\label{eq:wi_optimization_problem}
\begin{aligned}
\min_{w_i} &\quad \frac{1}{2}\norm{\Delta w_i}_{L^2(\Omega)}^2 \\
\text{such that} &\quad w_i(x_j) = \delta_{ij}, \quad j=1,\dots,q.
\end{aligned}
\end{equation}
Here $\Delta$ is the Laplacian on $\Omega$ with pure Neumann boundary conditions, and the right hand side of the constraint, $\delta_{ij}$, is the Kronecker delta, which takes value one if $i=j$ and zero if $i \neq j$. We refer to $\widetilde{\varphi}_x$ as the \emph{Poisson interpolant} of $\varphi_x$.

Let $u_y$ denote the function $u_y(x) := \widetilde{\varphi}_x(y)$ for an arbitrary but fixed value of $y$. It is straightforward to show that $u_y$ solves the optimization problem
\begin{equation}
\label{eq:uy_optimization_problem}
\begin{aligned}
\min_{u_y} &\quad \frac{1}{2}\norm{\Delta u_y}_{L^2(\Omega)}^2 \\
\text{such that} &\quad u_y(x_i) = \varphi_{x_i}(y), \quad i=1,\dots,q.
\end{aligned}
\end{equation}
This follows from linearity of the KKT system associated with \eqref{eq:wi_optimization_problem}, which in turn follows from the fact that the the objective function in \eqref{eq:wi_optimization_problem} is quadratic, and the constraint is linear.
Since the Laplacian measures local deviation from flatness [cite Jacobsen], from \eqref{eq:uy_optimization_problem} we see that the interpolant $x \mapsto u_y(x) = \widetilde{\varphi}_x(y)$ is the `flattest possible'' function (in a least-squares sense) that has Neumann boundary conditions and interpolates the data $\varphi_{x_i}(y)$.

Optimization problems \eqref{eq:wi_optimization_problem} and \eqref{eq:uy_optimization_problem} have unique continuous solutions when the spatial dimension is $1$, $2$, or $3$. When inverting for parameter fields defined on $4$ or higher spatial dimensions, the solution to this optimization problem does not exist, and Poisson interpolation (as described here) should not be used\footnote{The reason for this dependence on spatial dimension is related to the fact that the fundamental solution of the biharmonic equation is continuous in $3$ or less dimensions, but singular in $4$ or more dimensions. In the unusual scenario where one is inverting for a parameter field in $4$ or more spatial dimensions, one could use a modified Poisson interpolation where $\Delta$ is replaced with $\Delta^\alpha$ in the objective function of \eqref{eq:wi_optimization_problem}, where $\alpha>1$ is chosen so that the fundamental solution of $\Delta^{2\alpha}$ is continuous.}.

A key advantage of Poisson interpolation over other PSF interpolation methods, most notably radial basis functions [cite Escande etc], is that Poisson interpolation is \emph{geometry-aware}---Poisson interpolation incorporates information about the domain geometry when constructing the weighting functions. Consider the Pine island glacier geometry shown in Figure REF. Regions of the glacier are separated by thin inlets of ocean. With radial basis function interpolation, PSF samples on one side of the inlet inappropriately influence the interpolation on the other side of the inlet. With Poisson interpolation, the PSF samples only influence the interpolation on their own side of the inlet.

In Section SEC, we will show how to efficiently solve optimization problem \eqref{eq:wi_optimization_problem}, and thus compute $w_i$. Briefly, constructing all $q$ of the functions $w_i$ requires solving the Poisson equation $2q$ times with different right hand side sources, and these Poisson PDE solves can be performed efficiently with multigrid. Radial basis function interpolation of $\varphi_x(y)$ is cheaper per sample point because radial basis functions are given by analytic formulas. However, the cost of these Poisson PDE solves is typically cheap compared to the cost of computing the impulse response samples, $\varphi_{x_i}$. Computing the impulse response samples, $\varphi_{x_i}$, requires applying the Hessian to vectors, which in turn requires performing PDE solves of the linearized forward and ajoint equations for the inverse problem. These forward and adjoint PDE solves are typically much more expensive than Poisson PDE solves, because the Poisson equation is well-studied and easy to solve with multigrid.

% For example,
% \begin{itemize}
%     \item In inverse problems where the forward model is time-dependent, computing Hessian matrix-vector products requires solving time-dependent PDEs. In contrast, the Poisson equation is a stationary PDE.
%     \item In inverse problems where the parameter being inverted for resides on a boundary, computing Hessian matrix-vector products requires solving PDEs that are defined on a higher dimensional geometric domain than the Poisson PDEs that must be solved for the interpolation.
%     \item In multi-experiment inverse problems, data from many different ``experiments'' on the same medium are used in the inversion. Performing a Hessian matrix-vector product requires solving two PDEs \emph{per experiment}. The number of Poisson solves used to construct the weighting functions does not depend on the number of experiments.
% \end{itemize}
% The additional cost of Poisson interpolation is, therefore, a worthwhile investment if it allows one to achieve the same level of accuracy as other interpolation methods but with less sample points.

% One may also show that the weighting functions $w_i$ sum to one pointwise.

\subsection{Computation of Poisson weighting functions}

After discretization, optimization problem \eqref{eq:wi_optimization_problem} becomes
\begin{equation}
\label{eq:discrete_poisson_optimization_1}
\begin{aligned}
\min_{\mathbf{w}} &\quad \frac{1}{2}\mathbf{w}^T K^T M^{-1} K \mathbf{w} \\
\text{such that} &\quad E \mathbf{w} = \mathbf{d}.
\end{aligned}
\end{equation}
Here $\mathbf{w}$ is the discretization of $w_i$, $K$ is the stiffness matrix for the discretized Poisson problem with pure Neumann boundary conditions, $M$ is the mass matrix, and $\mathbf{d}=(0,\dots,0,1,0,\dots,0)$ is the length-$q$ vector with $i^\text{th}$ entry one and all other entries zero. Furthermore, $E$ is the pointwise observation matrix defined by $\left(E\mathbf{u}\right)_j = u(x_j)$, where $u$ denotes the function represented by the vector $\mathbf{u}$. Here we dropped the subscripts $i$ from variables for notational convenience. One must solve $q$ optimization problems of the form \eqref{eq:discrete_poisson_optimization_1}; one for each $i=1,\dots,q$.

In the remainder of this section we will derive an efficient algorithm for solving \eqref{eq:discrete_poisson_optimization_1}. The algorithm (Algorithm REF) is based on explicit formula for the solution of \eqref{eq:discrete_poisson_optimization_1} which we present in Proposition \ref{prop:poisson_interpolation_formula}. The formulas in Proposition \ref{prop:poisson_interpolation_formula} rely on certain matrices that we define in Definition \ref{defn:poisson_interpolation_matrices}. The primary computational cost of the algorithm is the solution of Poisson problems with pure Neumann boundary conditions.

\begin{defn}
	\label{defn:poisson_interpolation_matrices}
	Let 
	\begin{align*}
		\Psi :=& K^+ E^T, \\
		\Theta :=& K^+ M \Psi \\
		S :=& \Psi^T M \Psi,
	\end{align*}
	where $K^+$ is the Moore-Penrose pseudoinverse of $K$. Further, let $\mathbf{c}$ dente the discretization of the constant function $c(x)=1$ on $\Omega$, and let $\mathbf{1}:=(1,\dots,1)$ denote the vector of length $q$ with all entries equal to one.
\end{defn}

\begin{remark}
	Computing $\mathbf{g} = K^+\mathbf{f}$ is equivalent to solving the discretized Poisson problem with $f$ as the right hand side source, with pure Neumann boundary conditions, and with the constraint that the solution has average value equal to zero. That is,
	\begin{equation*}
		\mathbf{g} = K^+ \mathbf{f} \quad \Leftrightarrow \quad 
		\begin{cases}
			\Delta g = f, & \text{in }\Omega,\\
			\nu \cdot \nabla g = 0, & \text{on }\partial \Omega,\\
			\int_\Omega g(x) dx = 0,
		\end{cases}
	\end{equation*}
	where $\nu$ denotes the normal to the boundary.

	Hence, the $i^\text{th}$ column of $\Psi$ is the result of solving the discretized pure Neumann Poisson problem, with a point source at location $x_i$. The $i^\text{th}$ column of $\Theta$ is the result of solving the discretized pure Neumann Poisson problem with distributed source given by the $i^\text{th}$ column of $\Psi$.
\end{remark}

\begin{prop}
	\label{prop:poisson_interpolation_formula}
	The solution to optimization problem \eqref{eq:discrete_poisson_optimization_1} is given by
	\begin{equation*}
		\mathbf{w} = -\Theta \boldsymbol{\lambda} + \alpha \mathbf{c},
	\end{equation*}
	where
	\begin{align*}
		\alpha =& \left(\mathbf{1}^T S^{-1} \mathbf{1}\right)^{-1}\mathbf{1}^T S^{-1} \mathbf{d} \\
		\boldsymbol{\lambda} =& -S^{-1} \left(\mathbf{d} - \alpha \mathbf{1} \right). \\
	\end{align*}
\end{prop}

\begin{proof}
	Let
	\begin{equation*}
		\mathbf{v} := K \mathbf{w}.
	\end{equation*}
	Since the null-space of the Laplacian with pure Neumann boundary conditions is the set of all constant functions, we have
	\begin{equation}
	\label{eq:definition_of_w}
	\mathbf{w} = K^+ \mathbf{v} + \alpha \mathbf{c}
	\end{equation}
	for some scalar $\alpha$.

	Writing optimization problem \eqref{eq:discrete_poisson_optimization_1} in terms of $\mathbf{v}$ and $\alpha$ instead of $\mathbf{w}$ yields the following equivalent optimization problem:
	\begin{equation}
	\label{eq:discrete_poisson_optimization_2}
	\begin{aligned}
	\min_{\mathbf{v}, \alpha} &\quad \frac{1}{2}\mathbf{v}^T M^{-1} \mathbf{v} \\
	\text{such that} &\quad EK^+ \mathbf{v} + \alpha \mathbf{1} = \mathbf{d}.
	\end{aligned}
	\end{equation}
	The Lagrangian for optimization problem \eqref{eq:discrete_poisson_optimization_2} is
	\begin{equation*}
		\mathcal{L} = \frac{1}{2}\mathbf{v}^T M^{-1} \mathbf{v} + \boldsymbol{\lambda}^T\left(EK^+ \mathbf{v} + \alpha \mathbf{1} - \mathbf{d}\right).
	\end{equation*}
	The solution to this optimization problem is the stationary point of the Lagrangian, i.e., the point at which the gradient of the Lagrangian is zero:
	\begin{equation*}
		0 = \nabla \mathcal{L} = 
		\begin{bmatrix}
			\nabla_\mathbf{v} \mathcal{L} \\
			\nabla_{\boldsymbol{\lambda}} \mathcal{L} \\
			\nabla_\alpha \mathcal{L}
		\end{bmatrix} 
		= \begin{bmatrix}
			M^{-1} \mathbf{v} + K^+ E^T \boldsymbol{\lambda} \\
			EK^+\mathbf{v} + \alpha \mathbf{1} - \mathbf{d} \\
			\mathbf{1}^T \boldsymbol{\lambda}
		\end{bmatrix}.
	\end{equation*}
	This equation may be rewritten as the following block $3 \times 3$ linear system:
	\begin{equation*}
		\begin{bmatrix}
			M^{-1} & K^+ E^T & 0 \\
			EK^+ & 0 & \mathbf{1} \\
			0 & \mathbf{1}^T & 0
		\end{bmatrix}
		\begin{bmatrix}
			\mathbf{v} \\ \boldsymbol{\lambda} \\ \alpha
		\end{bmatrix}
		=
		\begin{bmatrix}
			0 \\ \mathbf{d} \\ 0
		\end{bmatrix}.
	\end{equation*}
	Performing block Gaussian elimination on this system allows us to reduce the system to the following block triangular form
	\begin{equation*}
		\begin{bmatrix}
			M^{-1} & K^+ E^T & 0 \\
			& -S & \mathbf{1} \\
			0 & 0 & \mathbf{1}^T S^{-1} \mathbf{1}
		\end{bmatrix}
		\begin{bmatrix}
			\mathbf{v} \\ \boldsymbol{\lambda} \\ \alpha
		\end{bmatrix}
		=
		\begin{bmatrix}
			0 \\ \mathbf{y} \\ \mathbf{1}^T S^{-1} y
		\end{bmatrix}
	\end{equation*}
	where we recall that $S := \Psi^T M \Psi = EK^+ M K^+ E^T$. From this block triangular system, we read off the solution as:
	\begin{align*}
		\alpha =& \left(\mathbf{1}^T S^{-1} \mathbf{1}\right)^{-1}\mathbf{1}^T S^{-1} \mathbf{d} \\
		\boldsymbol{\lambda} =& -S^{-1} \left(\mathbf{d} - \alpha \mathbf{1} \right) \\
		\mathbf{v} =& -M K^+ E^T \boldsymbol{\lambda}
	\end{align*}
	Substituting these results into the definition of $\mathbf{w}$ in \eqref{eq:definition_of_w}, we have
	\begin{equation*}
		\mathbf{w} = -K^+ M K^+ E^T \boldsymbol{\lambda} + \alpha \mathbf{c} = -\Theta \boldsymbol{\lambda} + \alpha \mathbf{c},
	\end{equation*}
	as required.
\end{proof}

We now design an incremental algorithm for constructing interpolants $\mathbf{u}$, in which we can add points $x_i$ more efficiently. Let
\begin{align*}
	\Psi :=& K^+ E^T \\
	\Theta :=& K^+ M \Psi
\end{align*}
so that 
\begin{equation*}
	S = \Psi^T M \Psi
\end{equation*}
and
\begin{equation*}
	\mathbf{u} = -\Theta \boldsymbol{\lambda} + \alpha \mathbf{c}
\end{equation*}
The columns of $\Psi$ are the responses of $A^+$ to point sources (delta distributions) centered at the sample points $x_i$. The columns of $\Theta$ are the responses of $A^+M^{-1} A$ to these point sources. The matrices $\Psi$ and $\Theta$ may be constructed incrementally as new points are added.

The solution to this system of equations is found by the following process:
\begin{enumerate}
	\item Compute 
	\begin{equation*}
		\Psi := A^+ B^T
	\end{equation*}
	by solving Neumann poisson problems for point source right hand sides, with point sources located at the points $x_i$, $i=1,\dots,q$.
	\item Compute 
	\begin{equation*}
		\Theta := A^+ M \Psi
	\end{equation*}
	by solving Neumann Poisson problems with the columns of $\Psi$ as distributed sources.
	\item Form 
	\begin{equation*}
		S = \Psi^T M \Psi
	\end{equation*}.
	\item Compute \begin{equation*}
		\alpha = \left(\mathbf{1}^T S^{-1} \mathbf{1}\right)^{-1}\mathbf{1}^T S^{-1} \mathbf{y}.
	\end{equation*}
	\item Compute 
	\begin{equation*}
		\boldsymbol{\lambda} = -S^{-1} \left(\mathbf{y} - \alpha \mathbf{1} \right).
	\end{equation*}
	\item Compute
	\begin{equation*}
		\mathbf{u} = -\Theta \boldsymbol{\lambda} + \alpha \mathbf{c}
	\end{equation*}
\end{enumerate}

\begin{thm}
	The weighting functions form a partition of unity.
\end{thm}

\section{Interpolating impulse response samples old}
